Collusion at the 1962 chess world championships

### 2019-05-02 inst/collusion_chess_draws.Rmd
more work

Now generate the dataset:

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, The Bradley-Terry model can be augmented
to accommodate draws in a number of ways.  Here I present a new
augmentation of Bradley-Terry in which a drawn chess match is regarded
as a three-way competition between the players and an entity that wins
if the match is drawn.  The resulting likelihood functions are easily
optimised numerically and I present a number of results.

Consider first the standard Bradley-Terry model for $n$ players.
First we ignore the effect of playing white.  We wish to find numbers
("strengths") $p_i,1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum
p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

$$
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
$$

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Incorporation of draws is
straightforward: we consider a chess match to be a three-way trial
between the two players and an entity that wins if the match is drawn.
We thus seek a likelihood function for non-negative
$p_1,\ldots,p_n,p_D$ $p_D+\sum p_i=1$ and this would be

\[
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\]

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by playing white.  We
thus consider a chess match between two players to be a three-way
entity but with a friendly ghost with strength $p_W$ whose strength is
added to the white player.

\[
\prod_{i<j}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\]

where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
playing white against $j$, and so on.  The final term is algebraically
equal to

\[
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
\]

which might be preferable on grounds of clarity.  We now apply this
likelihood model to a real dataset.
